A performance comparison of Dask and Spark for data-intensive neuroimaging pipelines

### 2019-07-25 latex/paper.tex
made fixes to Dask and the remainder of the methods section

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine.
While its core is written in Scala, Spark provides additional APIs in Java, R
and Python (PySpark). Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. This
data structure is the basis of Spark's other data structures, namely, DataFrames
and Datasets. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution to further improve performance. DataFrames, used to process
tabular data, Datasets where the data is organized into named-columns. While 
the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

\subsubsection{Dask} Dask is a Big Data engine that is becoming increasingly popular
in the scientific Python ecosystem. Dask has five main data structure APIs:
Array, Bag, DataFrame, Delayed, and Futures. All APIs were used in our experiments,
with the exception of Dask Dataframes, which is intended for use with tabular data.
Like Spark, all Dask APIs provide
in-memory computing, data locality, lazy evaluation, and fault tolerance by mean of
lineage. Dask's core is purely in Python and bypasses most of the
\href{https://docs.python.org/3/glossary.html#term-gil}{GIL} to provide parallelism.
We used Dask due to the common use of the scientific Python ecosystem in
neuroscience\VHS{If you add Dask Arrays to your experiments here, i'd mention that their
APIs might be particularly well suited for the processing of high resolution neuroimaging data}. Dask represents application tasks as Dask graphs, produced by a
user-facing API and executed by a scheduler. In Dask, operations generate multiple
tiny tasks in the computation graph allowing an easier representation of complex
algorithms. We used the \href{https://distributed.dask.org/en/latest/index.html}{Dask
Distributed} scheduler, although \href{https://github.com/dask/dask-yarn}{Dask-Yarn}
and \href{https://github.com/mrocklin/dask-mesos}{Dask-Mesos} is also available. In
the Dask Distributed scheduler, a process called \textit{dask-scheduler}
administrates the resources provided by \textit{workers} in the cluster. The
scheduler receives jobs from clients and assigns tasks to available workers. Like
Spark scheduler, Dask Distributed also use a LIFO policy, it is meant to reduce the
memory footprint of the application by finishing the computation of a branch in the
Dask graph before starting new ones. We used Dask v1.1.4.

In Dask, a \href{https://docs.dask.org/en/latest/bag.html}{Bag} is a parallel
collection of Python objects, similar to Spark's RDD. It use multi-process to enable
parallelism. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. An
\href{https://docs.dask.org/en/latest/array.html}{Array} is used for the processing
of large arrays. It provides a distributed clone of the popular NumPy library. A
\href{https://docs.dask.org/en/latest/dataframe.html}{Dataframe} is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas
Dataframes} used to process a large amount of tabular data.
\href{https://docs.dask.org/en/latest/delayed.html}{Delayed} supports arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
\href{https://docs.dask.org/en/latest/futures.html}{Futures} are similar to Delayed
as they support arbitrary tasks, but they operate in real-time rather than lazily.

@@

 We used Compute Canada's
 \href{https://docs.computecanada.ca/wiki/Cloud_resources}{Arbutus Cloud} operated by
 the \href{https://www.westgrid.ca}{WestGrid} regional organization at the University
 of Victoria, and running OpenStack \TG{version?}. We used c8-30gb-186 cloud
 instances with 8 VCPUs, an Intel Xeon Gold 6130 processor, \SI{30}{\giga\byte} of
 RAM at \SI{2666}{\mega\hertz}, \SI{20}{\giga\byte} of mounted storage, and a base
 image running CentOS 7.5.1804 with Linux kernel version
 3.10.0\-862.11.6.el7.x86\_64. Instances are connected by a
 \SI{10}{\giga\bit/\second} Ethernet network.

 Cloud instances hosted a single Dask or Spark worker, configured to use 8 CPUs. We
 used Dask's default configuration that uses all the available memory on the
 instance. Its default heuristic is to: target a 60\% memory load, spill to disk at
 70\%, pause the worker at 80\%, and terminate the worker at 95\%. We configured
 Spark to use 1 executor per worker and \SI{25}{\giga\byte} of memory per executor, to leave 5~GB
 for off-heap. We configured the Spark driver to use 25~GB of memory, and used the
 default configuration for the master. We used the default confirguration for worker
 memory management: at 60\% it spill data to disk, and 50\% of that amount is
 reserved for storage and is immuned from eviction.

 One cloud instance did not host any worker, and had a \SI{2}{\tera\byte} disk volume
 shared with the other instances using the Network File System (NFS) v4.
 This instance was also used for the Spark driver and master, and for the
 Dask scheduler, and for job monitoring with the Spark and Dask user
 interfaces. For both Spark and Dask, spilled data is evicted to the NFS.

@@

We used BigBrain~\cite{Amunts:13}, a three-dimensional image of a human
brain with voxel intensities ranging from 0 to 65,535. The original data is
stored in 125 blocks in the MINC~\cite{minc} HDF5-based format, available
at \url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D_Blocks/40um} at the
resolution of \SI{40}{\micro\metre}. We converted the blocks into the
\href{https://nifti.nimh.nih.gov/nifti-1}{NifTI} format, a popular format
in neuroimaging. We left the NifTI blocks uncompressed, resulting in 
a total data size of \SI{81}{\giga\byte}. 
To evaluate the effect of block size, we resplit these blocks into 30, 125 and 750 blocks of 
\SI{2.7}{\giga\byte}, \SI{0.648}{\giga\byte}, and
\SI{0.108}{\giga\byte}, using the sam~\cite{sam} library.

@@

We used three neuroimaging applications to evaluate the engines in different
conditions. The first two ones, incrementation and histogram, are simple synthetic
benchmarks representing basic map-only and map-reduce applications. The third one is
a realistic application representative of popular BIDS
applications~\cite{gorgolewski2017bids}. All script used for our experiment is availble
GitHub at \url{https://github.com/big-data-lab-team/paper-big-data-engines}

%% Incrementation
\subsubsection{Incrementation}
We used an adaptation of the simple image incrementation pipeline used
in~\cite{hayot2019performance} (see Algorithm~\ref{alg:incrementation}).
The application reads blocks of the BigBrain image from the shared file
system, increments the intensity value of each voxel by 1 to avoid caching
effects, sleeps for a configurable amount of time to emulate a more complex
processing, repeats this process for a specified amount of iterations, and
finally writes the result as a NifTI image back to the shared file system.
This application allows us to study the behavior of the engines when all
inputs are processed independently, in a map-only way (see
Figure~\ref{fig:tg-inc}). This mimics the behavior of analyzing multiple
independent subjects in parallel.

@@

 Our second application calculates the histogram of the BigBrain image (see
 Algorithm~\ref{alg:histogram}). It reads the image blocks from the shared file
 system, calculates intensity frequencies, aggregates the frequencies across the
 blocks, and finally writes the resulting histogram to the shared file system as a
 single \SI{766}{\kilo\byte} file. This application implements a typical map-reduce
 pattern, where the final result is obtained from all the input blocks (see
 Fig.~\ref{fig:tg-histo}). This application requires data shuffling thus inter-worker
 communication. The total amount of shuffled data is however limited to
 \SI{2.62}{\mega\byte} per blocks as it only consists of image histograms.

@@

We used the \href{https://github.com/BIDS-Apps/example}{BIDS app example} application
available on GitHub. This application operates in a map-reduce way. The map phase,
also called participant analysis, extracts the brain tissues of a subject's 3D MRI
using the popular FMRIB Software Library
(\href{https://fsl.fmrib.ox.ac.uk/fsl/fslwiki}{FSL}), and writes the resulting image
(\SI{2.47}{\mega\byte} on average) to the shared file system. The reduce phase, also
called group analysis, computes the volume of each brain and returns the average
volume, shuffling a total of \SI{8.6}{\giga\byte} image data.

>>>>

\subsubsection{Apache Spark} Apache Spark is a general-purpose Big Data engine.
While its core is written in Scala, Spark provides additional APIs in Java, R
and Python (PySpark). Its main abstraction, the Resilient Distributed Dataset
(RDD)~\cite{RDD}, is a fault-tolerant, parallel collection of data elements. It 
achieves fault tolerance through recording the data lineage, the sequence of 
operations required to produce the data. The RDD is the basis of Spark's other
data structures, namely, DataFrames
and Datasets. Datasets are similar to RDDs but additionally use Spark SQL's
optimized execution to further improve performance. DataFrames, used to process
tabular data, Datasets where the data is organized into named-columns. While 
the DataFrame API exists in all the available language APIs, Datasets are
limited to Scala and Java.

@@

\subsubsection{Dask} Dask is a Big Data engine that is becoming increasingly popular
in the scientific Python ecosystem. Like Spark, Dask provides
in-memory computing, data locality, lazy evaluation, and fault tolerance by
means of data lineage. In contrast, Dask's core is purely in Python, eliminating
the cost of serialization to Java. Furthermore, it is lightweight and modular,
allowing users to only install the components they require.

Dask has five main data structure APIs:
Array, Bag, DataFrame, Delayed, and Futures. A Dask
\href{https://docs.dask.org/en/latest/array.html}{Array} is used for the
processing of large arrays. It provides a distributed clone of the popular NumPy
library. Similar to Spark's RDD, a Dask
\href{https://docs.dask.org/en/latest/bag.html}{Bag} is a parallel
collection of Python objects. It offers a programming abstraction similar to the
\href{https://toolz.readthedocs.io/en/latest/}{PyToolz library}. A
\href{https://docs.dask.org/en/latest/dataframe.html}{Dataframe} is a parallel
composition of
\href{http://pandas.pydata.org/pandas-docs/stable/reference/api/pandas.DataFrame.html}{Pandas
Dataframes} used to process a large amount of tabular data.
\href{https://docs.dask.org/en/latest/delayed.html}{Delayed} supports arbitrary tasks
that do not fit in the Array, DataFrame or Bag APIs. Finally,
\href{https://docs.dask.org/en/latest/futures.html}{Futures} are similar to Delayed
as they support arbitrary tasks, but they operate in real-time rather than lazily.
All Dask data structure APIs, with the exception of Dask Arrays and Dask
DataFrames, were used in our experiments.

While Python multithreading is limited by the
\href{https://docs.python.org/3/glossary.html#term-gil}{Global Interpreter Lock (GIL)}
as a result of CPython's non thread-safe memory management, Dask attempts to
bypass the GIL whenever possible and use multithreading to provide parallelism.
With the exception of Dask Bags, all APIs, by default, use the local multithreaded
scheduler. Dask Bags instead relies on the local multiprocessing scheduler.

The Dask graph is the internal representation of Dask application to be executed 
by the scheduler. API Operations generate multiple
small tasks in the computation graph allowing an easier representation of complex
algorithms.

The Dask framework is compatible with numerous distributed schedulers, including 
\href{https://github.com/dask/dask-yarn}{YARN} and
\href{https://github.com/mrocklin/dask-mesos}{Mesos}, similarly to Spark. Dask 
also provides its own distributed scheduler, known as the Dask Distributed
scheduler. Although Dask is compatible with schedulers commonly found in HPC
clusters, we chose to use Dask's Distributed scheduler, to keep the environment
balanced between both frameworks.

 In the Dask Distributed scheduler, a process called \textit{dask-scheduler}
administrates the resources provided by \textit{workers} in the cluster. The
scheduler receives jobs from clients and assigns tasks to available workers. Similarly
to Spark's scheduler, Dask Distributed completes the processing of a graph branch
before moving along to the next one.

In our experiments, we used Dask v1.1.4.

@@

 We used Compute Canada's
 \href{https://docs.computecanada.ca/wiki/Cloud\_resources}{Arbutus Cloud} operated by
 the \href{https://www.westgrid.ca}{WestGrid} regional organization at the University
 of Victoria, running OpenStack Queens release. We used c8-30gb-186 cloud
 instances with 8 VCPUs, an Intel Xeon Gold 6130 processor, \SI{30}{\giga\byte} of
 RAM at \SI{2666}{\mega\hertz}, \SI{20}{\giga\byte} of mounted storage, and a base
 image running CentOS 7.5.1804 with Linux kernel version
 3.10.0\-862.11.6.el7.x86\_64. Instances are connected by a
 \SI{10}{\giga\bit/\second} Ethernet network.

 Cloud instances hosted a single Dask or Spark worker, configured to use 8 CPUs. We
 used Dask's default configuration that uses all the available memory on the
 instance. Its default heuristic is to: target a 60\% memory load, spill to disk at
 70\%, pause the worker at 80\%, and terminate the worker at 95\%. We configured
 Spark to use 1 executor per worker and \SI{25}{\giga\byte} of memory per executor, to leave 5~GB
 for off-heap. We configured the Spark driver to use 25~GB of memory, and used the
 default configuration for the master. We used the default confirguration for worker
 memory management: at 60\% it spill data to disk, and 50\% of that amount is
 reserved for storage and is immune to eviction.

 One cloud instance did not host any worker and had a \SI{2}{\tera\byte} disk volume
 shared with the other instances using the Network File System (NFS) v4.
 This instance was also used for the Spark driver and master, and for the
 Dask scheduler, and for job monitoring with the Spark and Dask user
 interfaces. For both Spark and Dask, spilled data is evicted to the NFS.

@@

We used BigBrain~\cite{Amunts:13}, a three-dimensional image of a human
brain with voxel intensities ranging from 0 to 65,535. The original data is
stored in 125 blocks in the MINC~\cite{minc} HDF5-based format, available
at \url{ftp://bigbrain.loris.ca/BigBrainRelease.2015/3D\_Blocks/40um} at a
resolution of \SI{40}{\micro\metre}. We converted the blocks into the
\href{https://nifti.nimh.nih.gov/nifti-1}{NifTI} format, a popular format
in neuroimaging. We left the NifTI blocks uncompressed, resulting in 
a total data size of \SI{81}{\giga\byte}. 
To evaluate the effect of block size, we resplit these blocks into 30, 125 and 750 blocks of 
\SI{2.7}{\giga\byte}, \SI{0.648}{\giga\byte}, and
\SI{0.108}{\giga\byte}, using the sam~\cite{sam} library.

@@

Three neuroimaging applications to evaluate the engines in different
conditions were used. The first two, incrementation and histogram, are simple 
synthetic applications representing map-only and map-reduce applications,
respectively. The third one is a real neuroimaging application representative
of popular BIDS applications~\cite{gorgolewski2017bids}. All scripts used for our
experiments are available at
\url{https://github.com/big-data-lab-team/paper-big-data-engines}

%% Incrementation
\subsubsection{Incrementation}
We used an adaptation of the image incrementation pipeline used
in~\cite{hayot2019performance} (see Algorithm~\ref{alg:incrementation}).
The application reads blocks of the BigBrain image from the shared file
system, increments the intensity value of each voxel by 1 to avoid caching
effects, sleeps for a configurable amount of time to emulate more
compute-intensive processing, repeats this process for a specified number of
iterations, and
finally writes the result as a NifTI image back to the shared file system.
This application allows us to study the behavior of the engines when all
inputs are processed independently, in a map-only scenario (see
Figure~\ref{fig:tg-inc}). This mimics the behavior of analyzing multiple
independent subjects in parallel.

@@

 Our second application calculates the histogram of the BigBrain image (see
 Algorithm~\ref{alg:histogram}). It reads the image blocks from the shared file
 system, calculates intensity frequencies, aggregates the frequencies across the
 blocks, and finally writes the resulting histogram to the shared file system as a
 single \SI{766}{\kilo\byte} file. This application implements a typical map-reduce
 program, where the final result is obtained from all the input blocks (see
 Fig.~\ref{fig:tg-histo}). This application requires data shuffling, incurring 
 additional performance costs compared to Algorithm~\ref{alg:incrementation}.
 The total amount of shuffled data is, however, limited to
 \SI{2.62}{\mega\byte} per blocks as it only consists of image histograms.

@@

We used the \href{https://github.com/BIDS-Apps/example}{BIDS app example} application
available on GitHub. This application can also be broken down into a map and
reduce phase. The map phase,
also called participant analysis, extracts the brain tissues of a subject's 3D MRI
using the popular FMRIB Software Library
(\href{https://fsl.fmrib.ox.ac.uk/fsl/fslwiki}{FSL}), and writes the resulting image
(\SI{2.47}{\mega\byte} on average) to the shared file system. The reduce phase, also
called group analysis, computes the volume of each brain and returns the average
volume, shuffling a total of \SI{8.6}{\giga\byte} image data.
